The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-22 paper.qmd
rewrote a lot of "uncertainty as a variable stuff"

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
fig-valign: bottom
editor_options: 
  chunk_output_type: console
---

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today. For example, the '95% confidence interval' is more accurately a '95% prediction interval'. Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty. So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information. The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-ncol: 3
#| fig-cap-location: "margin"

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. This trend is only technically there in the low variance condition, i.e. if we performed a hypothesis test we would conclude the trend exists. The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "Choropleth Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. This disagreement comes from two different conceptualisations of what an uncertainty visualisation *is*. There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate. The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.

@@

# Visualising Uncertainty as a Variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. You may think "well, I will just add a dimension to my plot that includes uncertainty". This is not a silly choice, but lets look at what happens when we do this.

## Bivariate Map
## What makes uncertainty visualisation difficult?

The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.

It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*. A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic. Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, median, maximum, or of any other statistic, however this field *does* implicitly treat them different. There is no such thing as "maximum visualiation" or "median visualisation" so there *must* be something about uncertainty that causes authors to believe it is special. Here we seek to answer this question.

More accurate to class this as "why is uncertainty visualisation difficult"
Uncertainty visualisation papers usually provide one of two primary justifications for their existence. The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty. Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019]. The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate. Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].

Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear. The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].

> Do we treat uncertainty as ‘just another variable’ to be visually represented or does it need to be treated differently? For example, when we picture a map showing air pressure distribution in combination with temperature, these two variables are certainly dependent on each other (in physical terms). The same is true with air pressure and its uncertainty, but we see a stronger dependency: Uncertainty can be seen as metadata of air- pressure which we argue makes a difference. Thus, we support Edwards and Nelson (2001, p. 35) who stated that ‘[p]erhaps data certainty information is unique and will require a new type of framework for designing symboliza- tion’. Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty. Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time. This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.

# Explicitly Combining Uncertainty and Signal
## Using Statistics
### What is an uncertainty visualisation?
## Using Aesthetics
### Value Supressing Uncertainty Palettes
### There is no such thing as a test for everything

# Implicitly Combining Uncertainty and Signal
## Pixel maps
## Just visualise your data

@@

### Direct questions
@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).

@@

# ------------------- BELOW HERE IS UNSORTED ------------------------
# Why is uncertainty special?

Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to. These experiments inherently view uncertainty as a variable that is separate to the variable on which we have mapped our signal. For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature. It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables. If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor. This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].

Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty. Hence the name of the pallet. @fig-maps depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal". The first and most basic map is a simple choropleth map where each value (the colouring of each local government area) has no associated "uncertainty". The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue. If visualisation was performing signal suppression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".  The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain. @Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness"). That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesn't occur in a way that is useful for uncertainty visualisation). Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted. It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for). Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable. This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-maps
#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map. There is a high variance and low variance example for each type of map to show how well the technique supresses signal. Each map was created using the same data with the same base palette. At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty. This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate. "

@@

# Display plots and palettes
p2a
p2b
show_col(bivariatepal, ncol=8, labels=FALSE)
p3a
p3b
show_col(VSUP, ncol=8, labels=FALSE)
p4a
p4b
show_col(basecols, ncol=8, labels=FALSE)
```

# What even is an uncertainty visualisation?

### What is an uncertainty visualisation?
@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy. It can be seen that most of the categories can be considered a change to the typical error bar chart. The taxonomy categorises the visualisations based on five criteria, implicit (a sample) or explicit (a depiction of mass); Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty); visually integral or separable (the uncertainty can be separated from the data and read independently) ; coincidence (uncertainty and data are represented in the same plot) or adjacent; static or dynamic (animated, interactive etc). Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-tax}

@@

Uncertainty is a concept that is famously hard to pin down, with constantly shifting definitions that mean it is quantified differently from author to author.

@@

Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific. @Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question. @Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants. Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in @fig-right. The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice. These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot. Therefore, this method leads to inconclusive results about the plot design, and is not advised.

![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation.](furthestright.png){#fig-right width=50%}

@@

![The visualisation showing the kind of signal that we would expect an uncertainty plot to supress](signalsupression.jpeg){#fig-sigsupress}

Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.

![A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference). This can be considered the graphical equivalent of a standard hypothesis test. Uncertainty is frequently used to descibe the area around the estimate that we think might contain the true value, and we assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). Uncertainty is technically the distance between the true value and the estimate. ](hypvs.jpeg){#fig-hypvs}

@@

>>>>

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today. For example, the '95% confidence interval' is more accurately a '95% prediction interval'. Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty. So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information. The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-ncol: 3

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. This trend is only technically there in the low variance condition, i.e. if we performed a hypothesis test we would conclude the trend exists. The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "Choropleth Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"
#| cap-location: "bottom"

@@

This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. We belive this disagreement comes from two different conceptualisations of what an uncertainty  visualisation *is*. This issue causing confusion in the literature was also commented on by @Kinkeldey2014, who explained it was unclear if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014]. There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate. The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.

@@

# Including uncertainty as an indepdendent variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. You may think "well, I will just add a dimension to my plot that includes uncertainty". This makes sense, as this is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel.

@fig-bivariate a variation of the choropleth map, where we have a two dimensional colour palette. Not only is temperature mapped to hue, but variance is also mapped to saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist). The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a uncertainty visualisation? And if not, why?

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-bivariate
#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature. This map is a choropleth map with a two dimensional colour palette where temperature is represented by colour hue, and variance is represented by colour saturation. Even though uncertainty has been added to the graphic the spatial trend is still clearly visible in the case where the spatial trend could be attributed to noise."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "Bivariate Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

show_pal2 <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, myylab, breaks, breaks2) {
  # Set dimensions of palette
  n <- length(colours)
  ncol <- ncol %||% ceiling(sqrt(length(colours)))
  nrow <- ceiling(n/ncol)
  # make matrix with null values (if not full)
  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
  # set graphical parameters (?)
  old <- par(pty = "s", mar = c(0, 0, 0, 0))
  on.exit(par(old))
  size <- max(dim(colours))
  plot(c(-1.5, size), c(0, -size), type = "n", xlab = "", ylab = "", 
      axes = FALSE)
  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
       col = colours, border = borders)
  text(c(0,col(colours)[nrow,]) + c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2) , -4.5, 
       breaks, cex = 1, col = "black")
  text(-0.25, -c(0,row(colours)[,ncol]) + c(-0.2, -0.1, 0, 0.1, 0.2),
       breaks2, cex = 1, col = "black")
  text(4, -5.5, myxlab ,cex = 1.5, col = "black")
  text(x=-1.25,y=-2, myylab, srt=270, cex = 1.5, col = "black")
}

p2a
p2b
show_pal2(colours = bivariatepal, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
```

## What makes uncertainty visualisation difficult?
Uncertainty visualisation is a uniquely difficult problem in visualisation. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty. There are two primary theories as to *why* uncertainty visualisation is unique.

The first group believes uncertainty visualisation are unique due to the psychological heuristics involved in interpreting uncertainty [@Spiegelhalter2017; @Hullman2019]. While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to a problem in visualisation. Were this the case, visualisations of death, poverty, and any other variable that elicits poor feelings would require its own field of visualisation.

The second group sees uncertainty as a high dimensional visualisation problem as the difficulty is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006]. If we view uncertainty visualisation as a high dimension problem we include uncertainty as we would any other variable, and allow for the signal to be interpreted independent of the variance. This prevents uncertainty and signal being read together as a "single integrated uncertain value" [@Kinkeldey2014] and allows for uncertainty to be ignored [@uncertchap2022]. Additioanlly, this integration needs to be communicated through the "gist" of the graphic, as that is primary purpose of a visualisation [@Spiegelhalter2017]. 
The @fig-bivariate plot's "gist" is "this data has a spatial trend and the estimates have a large variance". It is clear the uncertainty and signal are being communicated separately.

Uncertainty visualisation is difficult because we need the uncertainty axis to interfere with the signal axis. When we typically visualise two independent variables we *do not* want them to interfere with each other, however the opposite is true for uncertainty visualisation. For uncertainty to perform signal suppression, it must interfere with the visual channel that signal was mapped to. It needs to interfere enough that viewers do not see signals that are simply the product of noise.

Lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature. It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables. If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor.

 *** up to here ***

The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue. If visualisation was performing signal suppression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".

The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain. @Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness"). That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesn't occur in a way that is useful for uncertainty visualisation).

- seperability experiments (susan, animation, scatter size)
- suprise map (I think they had something about comparing light and dark)
- Something else on accessibility or colour space (check susan's work)

Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted. It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for). Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable. This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].

# Visualising uncertainty and signal in a new space
This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].

Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty. Hence the name of the pallet. depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal".

### Value Supressing Uncertainty Palettes
### There is no such thing as a test for everything

# Implicitly Combining Uncertainty and Signal
## Pixel maps
## Show me the data

@@

### Direct questions

> Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty. Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time. This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value. [@Kinkeldey2014]

@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).

@@

# ------------------- BELOW HERE IS UNSORTED ------------------------

# MAPS

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false

@@

# Display plots and palettes
p3a
p3b
show_col(VSUP, ncol=8, labels=FALSE)
p4a
p4b
show_col(basecols, ncol=8, labels=FALSE)
```

# What is an uncertainty visualisation?
@Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

@@

@@

Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific. @Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question. @Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants. Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in fig-right. The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice. These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot. Therefore, this method leads to inconclusive results about the plot design, and is not advised.

@@

Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.

@@

## Transparneyc map
```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
#| label: fig-transparency
#| fig-cap: ""
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "Choropleth Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

trans_map <- my_map_data |>
  mutate(scaled_var = (variance - min(variance))/(max(variance)-min(variance)))

# Transparency map
pta <- trans_map |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf(aes(fill = bitemp, 
              geometry = geometry,
              alpha=scaled_var)) + 
  scale_fill_manual(values = basecols) +
  theme_void() + 
  theme(legend.position = "none")

ptb <- pta %+% filter(trans_map, variance_class=="highvar")
